---
title: "01_viz"
author: "Emma"
date: "2026-10-01"
output: github_document
---

```{r message = F}
library(tidyverse)
library(ggridges)
```

```{r}
library(p8105.datasets)
data("weather_df")
```

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax))
```
This creates a blank scatterplot, no geom specified 

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point()
```

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()
```
Better way to write code for the same scatterplot

```{r}
ggp_weather = 
  weather_df |>
  ggplot(aes(x = tmin, y = tmax)) 

ggp_weather + geom_point()
```
Saving plot, not Jeff's default for making plots 

# Advanced Scatterplot
```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name))
```
Adding color based on name variable 

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) +
  geom_smooth(se = FALSE)
```
Adding curve and making dots slightly transparent

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

```{r}
weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```
Looking at times of year and precipitation 

```{r}
weather_df |>  
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_smooth(se = FALSE) 
```

```{r}
weather_df |> 
  ggplot(aes(x = tmax, y = tmin)) + 
  geom_hex()
```

# Learning assessment - do this 

# Univariate Plots 
## Histogram 
```{r}
weather_df |> 
  ggplot(aes(x = tmax)) + 
  geom_histogram()
```

Adding color:
```{r}
weather_df |> 
  ggplot(aes(x = tmax, fill = name)) + 
  geom_histogram(position = "dodge", binwidth = 2)
```

Kind of hard to understand, so make density plots instead:
```{r}
weather_df |> 
  ggplot(aes(x = tmax, fill = name)) + 
  geom_density(alpha = .4, adjust = .5, color = "blue")
```

## Box Plots 

```{r}
weather_df |> 
  ggplot(aes(y = tmax)) +
  geom_boxplot()
```
Showing distribution of temperatures 

```{r}
weather_df |> 
  ggplot(aes(x = name, y = tmax)) +
  geom_boxplot()
```
Add name variable on x axis, do three separate plots for each of the name variables, compare 

## Violin Plots 

```{r}
weather_df |> 
  ggplot(aes(x = name, y = tmax)) +
  geom_violin()
```
Same as box plot, but now violin. Helpful for some distribution types, eg bimodal distributions are hidden by box plots 

## Ridge Plots 

```{r}
weather_df |> 
  ggplot(aes(x = tmax)) +
  geom_density()
```
This is overall density of everything 

```{r}
weather_df |> 
  ggplot(aes(x = tmax, y = name)) +
  geom_density_ridges()
```
Density plots, but using ridges to separate out in vertical direction. Helpful to see what is going on if you have a lot of categories (because in density plot they all overlap)


