---
title: "02_viz"
author: "Emma"
date: "2026-10-06"
output: github_document
---
# Lecture Notes 

How data organization and manipulation feeds into making plots. A good picture is worth 1000 words! 
* Critical to look at datasets before making figures 

To have a good picture:

* Show as much of the data as you can 
* avoid superfluous frills (eg. 3D)
* Facilitate comparisons 
* Put groups in sensible order 
* use common axes 
* use a color to highlight groups 
* no pie charts (make a bar chart instead)
  
Good pictures are not necessarily publication quality. Most figures are for you, and these should still be good. Publication figures take much more time. 

Using ggplot: 
Basic graph components 

* data, aesthetic mappings, geoms 

Advanced graph components 

* facets, statistics, scales 

A graph is built by combining these components 
Graphics can be further customized depending on the goals 

* Axis labels, axis tick locations/labels, font sizes, graphs themes, color scales, combining panels 

Based around the "tidy data" framework 
Trouble making a plot is often trouble with data tidiness in disguise 

* think about your data organization and how it affects your ability to visualize
* Factors can help with ordering 

# Time to make some more complicated plots! 

```{r}
library(tidyverse)
library(p8105.datasets)

data("weather_df") 
```


```{r}
weather_df |> 
  ggplot(aes(x = tmax, y = tmin, color = name)) + 
  geom_point() +
  labs(
    title = "Temperature (Max vs. Min)", 
    x = "Max Temperature (C)", 
    y = "Min Temperature (C)", 
    color = "Location",
    caption = "Data from NOAA for three weather stations."
  )
```
Labeling axes, adding captions, etc. Jeff does not bother with labels in graphs for himself, just for other people. Still helpful. 

Let's try some other scales. 
```{r}
weather_df |> 
  ggplot(aes(x = tmax, y = tmin, color = name)) + 
  geom_point() +
  labs(
    title = "Temperature (Max vs. Min)", 
    x = "Max Temperature (C)", 
    y = "Min Temperature (C)", 
    color = "Location",
    caption = "Data from NOAA for three weather stations."
  ) + 
  scale_x_continuous(
    breaks = c (-10, 0, 15), 
    labels = c("-10 C", "0", "Fifteen")
  ) + 
  scale_y_continuous(
    trans = "sqrt",
    position = "right"
  )
```
Can control scale depending on the variable. different scale functions. This changes the scale on the x axis. You can also label the scales on the axes. Can also do transformations (square root), and change position of axis labels (eg. y axis label from left side to right side). 

Let's look at color! Jeff uses this scale all the time. 
```{r}
weather_df |> 
  ggplot(aes(x = tmax, y = tmin, color = name)) +
  geom_point() +
  scale_color_hue(h = c(100, 300)) 
```

use the viridis color palette for pretty much everything: 
```{r}
weather_df |> 
  ggplot(aes(x = tmax, y = tmin, color = name)) +
  geom_point() +
  viridis::scale_color_viridis(
    name = "Location",
    discrete = TRUE
  )
```
Have to say whether variable is discrete or continuous 



